|

### 2020-07-28 article_format/article_format_anon.Rmd
Changed the formating, made some minor edits

---
title: |
  A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}
#type: ARTICLE TEMPLATE
author:
  - name: Henrik Kenneth Andersen
    affil: a
    email: henrik.andersen@soziologie.tu-chemnitz.de
affiliation:
  - num: a
    address: |
      Chemnitz University of Technology, Institute of Sociology, Chair for Empirical Social Research, Thüringer Weg 9, 09126 Chemnitz, Germany
bibliography: references2.bib
# appendix: appendix.tex
abstract: |
  This article provides an in-depth look at fixed effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed effects in the `lavaan` package for `R`. It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output. Online supplementary materials further discuss various common extensions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/ random effects model.
keywords: |
  Fixed effects, structural equation modeling, lavaan, R, panel analysis
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{booktabs}
  \usepackage{bm}
  \usepackage{mathtools}
  \usepackage{amssymb}
  \usepackage{amsmath}
  \usepackage{tikz}
  \usetikzlibrary{arrows}
  \usepackage[nofiglist]{endfloat}
  \usepackage{blkarray}
  \usepackage{setspace}
  \usepackage{etoolbox}
  \DeclareMathOperator{\E}{\mathbb{E}}
  \DeclareMathOperator{\Var}{\mathrm{Var}}
  \DeclareMathOperator{\Cov}{\mathrm{Cov}}
  \DeclareMathOperator*{\argmax}{arg\,max}
  \DeclareMathOperator*{\argmin}{arg\,min}
  \mathtoolsset{showonlyrefs}
  \BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
  \BeforeBeginEnvironment{align}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}
  \def\tightlist{}
  \pagenumbering{gobble}
output: rticles::tf_article
---

>>>>

---
title: |
  A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}
#type: ARTICLE TEMPLATE
author:
  - name: ""
    affil: a
    email: ""
affiliation:
  - num: a
    address: |
bibliography: references2.bib
# appendix: appendix.tex
abstract: |
  This article provides an in-depth look at fixed effects regression in the structural equation modeling (SEM) framework, specifically the application of fixed effects in the `lavaan` package for `R`. It is meant as a applied guide for researchers, covering the underlying model specification, syntax, and summary output. Online supplementary materials further discuss various common extensions to the basic fixed-effect model, demonstrating how to relax model assumptions, deal with measurement error in both the dependent and independent variables, and include time-invariant predictors in a type of hybrid fixed-/ random effects model.
keywords: |
  Fixed effects, structural equation modeling, lavaan, R, panel analysis
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{booktabs}
  \usepackage{bm}
  \usepackage{mathtools}
  \usepackage{amssymb}
  \usepackage{amsmath}
  \usepackage{tikz}
  \usetikzlibrary{arrows}
  \usepackage[nofiglist]{endfloat}
  \usepackage{blkarray}
  \usepackage{setspace}
  \usepackage{etoolbox}
  \DeclareMathOperator{\E}{\mathbb{E}}
  \DeclareMathOperator{\Var}{\mathrm{Var}}
  \DeclareMathOperator{\Cov}{\mathrm{Cov}}
  \DeclareMathOperator*{\argmax}{arg\,max}
  \DeclareMathOperator*{\argmin}{arg\,min}
  \mathtoolsset{showonlyrefs}
  \BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
  \BeforeBeginEnvironment{align}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}
  \def\tightlist{}
  \pagenumbering{gobble}
output: rticles::tf_article
keep_tex: true
---

### 2020-07-28 article_format/extensions_format.Rmd
Changed the formating, made some minor edits

---
title: "Supplememtary materials for:"
#author: "Henrik Kenneth Andersen"
#date: "`r Sys.Date()`"
geometry: margin=1in
output: 
  pdf_document:
    fig_caption: yes
    number_sections: yes
    includes: 
      in_header: header.tex
documentclass: article
classoption: a4paper
urlcolor: blue
link-citations: yes
bibliography: references.bib   
fontsize: 12pt
---

\vspace{-75pt}

\pagenumbering{arabic} 
\setcounter{page}{1}

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# ----- Packages

library( formatR)
library( knitr)
library( lavaan)

source("scripthooks.R")
```

[Andersen, Henrik K. 2020. "A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}"](https://github.com/henrik-andersen/FE-SEM/blob/master/article.pdf).

```{r echo=FALSE}
df <- readRDS("longData.Rda")
dfw <- readRDS("wideData.Rda")

n <- readRDS("n.Rda")
waves <- readRDS("waves.Rda")

fe_sem <- readRDS("fe_sem.Rda")
fe_sem.fit <- readRDS("fe_sem.fit.Rda")
```

@@

From this we see both models report the same estimated variance components, $\hat{\sigma}^{2}_{\alpha} = `r round(ercomp(fe2)[[1]][2], 3)`$ and $\hat{\sigma}^{2}_{\varepsilon} = `r round(ercomp(fe2)[[1]][1], 3)`$, telling us that about `r round(ercomp(fe2)[[1]][2]/(ercomp(fe2)[[1]][1] + ercomp(fe2)[[1]][2]), 3)`% of the residual variance is due to the differences between individuals (shown in the `share` column of the `ercomp()` output). This is what is referred to as the intraclass correlation coefficient, or ICC [@Hox2010].

@@

Indeed, the magnitude of the bias in a multivariate model is somewhat more complex to derive, but it will be 
\begin{align}
\hat{\beta} & = \beta \frac{\Var(\theta)}{\Var(\theta) + \Var(\upsilon)}
\end{align}
where $\theta$ is just the residual of a regression in which the true underlying variable is regressed on all other covariates. In this case, we need to regress $x_{t}$ on $\alpha_{1}$ and $\alpha_{2}$ for: $x_{t} = \gamma_{1}\alpha_{1} + \gamma_{2}\alpha_{2} + \theta_{t}$ [@Wooldridge2009, p. 318--320].

Normally it is not possible to reconstruct the bias since in cases where we have to rely on indicators, we would not know the true underlying variable. Furthermore, in the case of a fixed-effects model, the covariates are the unobserved time-invariant characteristics. However, because we are working with simulated data, we have everything we need. Going back to the results above, we can get the residuals of $x_{t}$ by either running a regression and saving the residuals, or we could skip a step and get them directly using the 'residual maker' matrix [@Ruettenauer2020] which is $\bm{M} = \bm{I} - \bm{A}(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}$ and $\bm{A} = \begin{pmatrix}\bm{\alpha_{1}}, \bm{\alpha_{2}}\end{pmatrix}$ is the $n \times 2$ matrix of the covariates.

@@

\newpage

# References {-}

<div id="refs"></div>

>>>>

---
title: |
  | Supplementary materials for: 
  | A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}
#type: ARTICLE TEMPLATE
author:
  - name: Henrik Kenneth Andersen
    affil: a
    email: henrik.andersen@soziologie.tu-chemnitz.de
affiliation:
  - num: a
    address: |
      Chemnitz University of Technology, Institute of Sociology, Chair for Empirical Social Research, Thüringer Weg 9, 09126 Chemnitz, Germany
bibliography: references2.bib
# appendix: appendix.tex
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{booktabs}
  \usepackage{bm}
  \usepackage{mathtools}
  \usepackage{amssymb}
  \usepackage{amsmath}
  \usepackage{tikz}
  \usetikzlibrary{arrows}
  \usepackage[nofiglist]{endfloat}
  \usepackage{blkarray}
  \usepackage{setspace}
  \usepackage{etoolbox}
  \DeclareMathOperator{\E}{\mathbb{E}}
  \DeclareMathOperator{\Var}{\mathrm{Var}}
  \DeclareMathOperator{\Cov}{\mathrm{Cov}}
  \DeclareMathOperator*{\argmax}{arg\,max}
  \DeclareMathOperator*{\argmin}{arg\,min}
  \mathtoolsset{showonlyrefs}
  \BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
  \BeforeBeginEnvironment{align}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}
  \def\tightlist{}
  \pagenumbering{gobble}
output: rticles::tf_article
---

```{r setup, include=FALSE}
# Knitr options 
knitr::opts_chunk$set(echo = TRUE)

# Packages (some are repeated below, add them here as well just to be sure)
library(formatR)
library(knitr)
library(lavaan)

# Script hooks for chunk outputs
source("../scripthooks.R")
```

```{r echo=FALSE}
df <- readRDS("../longData.Rda")
dfw <- readRDS("../wideData.Rda")

n <- readRDS("../n.Rda")
waves <- readRDS("../waves.Rda")

fe_sem <- readRDS("../fe_sem.Rda")
fe_sem.fit <- readRDS("../fe_sem.fit.Rda")
```

@@

From this we see both models report the same estimated variance components, $\hat{\sigma}^{2}_{\alpha} = `r round(ercomp(fe2)[[1]][2], 3)`$ and $\hat{\sigma}^{2}_{\varepsilon} = `r round(ercomp(fe2)[[1]][1], 3)`$, telling us that about `r round(ercomp(fe2)[[1]][2]/(ercomp(fe2)[[1]][1] + ercomp(fe2)[[1]][2]), 3)*100`% of the residual variance is due to the differences between individuals (shown in the `share` column of the `ercomp()` output). This is what is referred to as the intraclass correlation coefficient, or ICC [@Hox2010].

@@

Indeed, the magnitude of the bias in a multivariate model is somewhat more complex to derive, but it will be 
\begin{align}
\hat{\beta} & = \beta \frac{\Var(\theta)}{\Var(\theta) + \Var(\upsilon)}
\end{align}
where $\theta$ is just the residual of a regression in which the true underlying variable is regressed on all other covariates. In this case, we need to regress $x_{t}$ on $\alpha_{1}$ and $\alpha_{2}$ for: $x_{t} = \tau + \gamma_{1}\alpha_{1} + \gamma_{2}\alpha_{2} + \theta_{t}$ where $\tau$ is the intercept, and $\gamma_{1}$, $\gamma_{2}$ are the regression coefficients and $\theta_{t}$ is the residual [@Wooldridge2009, p. 318--320].

Normally it is not possible to reconstruct the bias since in cases where we have to rely on indicators, we would not know the true underlying variable. Furthermore, in the case of a fixed-effects model, the covariates are the unobserved time-invariant characteristics. However, because we are working with simulated data, we have everything we need. Going back to the results above, we can get the residuals of $x_{t}$ by either running a regression and saving the residuals, or we could skip a step and get them directly using the 'residual maker' matrix [@Ruettenauer2020] which is $\bm{M} = \bm{I} - \bm{A}(\bm{A}^{\intercal}\bm{A})^{-1}\bm{A}^{\intercal}$ and $\bm{A} = \begin{pmatrix}\bm{\iota}_{n} & \bm{\alpha_{1}} & \bm{\alpha_{2}}\end{pmatrix}$ is the $n \times 3$ matrix of the covariates (plus a constant).

@@

# References

### 2020-07-28 article_format/extensions_format_anon.Rmd
Changed the formating, made some minor edits

---
title: |
  | Supplementary materials for: 
  | A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}
#type: ARTICLE TEMPLATE
author:
  - name: Henrik Kenneth Andersen
    affil: a
    email: henrik.andersen@soziologie.tu-chemnitz.de
affiliation:
  - num: a
    address: |
      Chemnitz University of Technology, Institute of Sociology, Chair for Empirical Social Research, Thüringer Weg 9, 09126 Chemnitz, Germany
bibliography: references2.bib
# appendix: appendix.tex
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{booktabs}
  \usepackage{bm}
  \usepackage{mathtools}
  \usepackage{amssymb}
  \usepackage{amsmath}
  \usepackage{tikz}
  \usetikzlibrary{arrows}
  \usepackage[nofiglist]{endfloat}
  \usepackage{blkarray}
  \usepackage{setspace}
  \usepackage{etoolbox}
  \DeclareMathOperator{\E}{\mathbb{E}}
  \DeclareMathOperator{\Var}{\mathrm{Var}}
  \DeclareMathOperator{\Cov}{\mathrm{Cov}}
  \DeclareMathOperator*{\argmax}{arg\,max}
  \DeclareMathOperator*{\argmin}{arg\,min}
  \mathtoolsset{showonlyrefs}
  \BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
  \BeforeBeginEnvironment{align}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}
  \def\tightlist{}
  \pagenumbering{gobble}
output: rticles::tf_article
---

>>>>

---
title: |
  | Supplementary materials for: 
  | A closer look at fixed effects regression in structural equation modeling using \texttt{lavaan}
#type: ARTICLE TEMPLATE
author:
  - name: ""
    affil: a
    email: ""
affiliation:
  - num: a
    address: |
bibliography: references2.bib
# appendix: appendix.tex
header-includes: |
  \usepackage{hyperref}
  \usepackage[utf8]{inputenc}
  \usepackage{booktabs}
  \usepackage{bm}
  \usepackage{mathtools}
  \usepackage{amssymb}
  \usepackage{amsmath}
  \usepackage{tikz}
  \usetikzlibrary{arrows}
  \usepackage[nofiglist]{endfloat}
  \usepackage{blkarray}
  \usepackage{setspace}
  \usepackage{etoolbox}
  \DeclareMathOperator{\E}{\mathbb{E}}
  \DeclareMathOperator{\Var}{\mathrm{Var}}
  \DeclareMathOperator{\Cov}{\mathrm{Cov}}
  \DeclareMathOperator*{\argmax}{arg\,max}
  \DeclareMathOperator*{\argmin}{arg\,min}
  \mathtoolsset{showonlyrefs}
  \BeforeBeginEnvironment{equation}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{equation}{\end{singlespace}\noindent\ignorespaces}
  \BeforeBeginEnvironment{align}{\begin{singlespace}\vspace*{-\baselineskip}}
  \AfterEndEnvironment{align}{\end{singlespace}\noindent\ignorespaces}
  \def\tightlist{}
  \pagenumbering{gobble}
output: rticles::tf_article
keep_tex: true
---

\pagenumbering{arabic} 
\setcounter{page}{1}
